{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-07-13 sftrends_manu.Rnw
c10 d7 trends plot

\subsection{Case studies}
These are science questions that are relevant outside of the region.
\subsubsection{Disaggregating observed nitrogen time series}
Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear and complex, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
\subsubsection{Effects of wastewater treatment}
Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions. See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
\subsubsection{Effects of biological invasions}
Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.

@@

\subsubsection{Clam invasion in Suisun Bay}
Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.  Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.

@@

@@

@@

cap.val<-'Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods. Summaries are means and percent changes based on annual means within the time periods.  Increasing values are in bold-italics.'

>>>>

\subsection{Case studies}
These are science questions that are relevant outside of the region.
\subsubsection{Disaggregating observed nitrogen time series}
Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear and complex, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
\subsubsection{Effects of wastewater treatment}
Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions. See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
\subsubsection{Effects of biological invasions}
Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.

@@

\subsubsection{Clam invasion in Suisun Bay}

Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.  Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.

@@

What do nitrogen trends mean?  Have to interpret relative to trends in other variables.  A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.  A change in chlorohpyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.  However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.  Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on? See Senn slide 14 (from burial?). The WRTDS model lets us at least address trends in the context of season, time, and flow.  This allows for more improved interpretion relative to observing raw data. Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN. What about other variables (light level as suspended particulate matter, temperature)?

@@

<<d7c10trnds, fig.cap = 'Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\\it\\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).  Covariation between nutrients, chl-{\\it\\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.  Seasonal changes at D7 are most pronounced during the summer.', fig.width = 9, fig.height = 10>>=
data(h3dat)

# separate each site
toplo1 <- filter(h3dat, Site_Code %in% 'D7')
toplo2 <- filter(h3dat, Site_Code %in% 'C10')

# y axis labels
ylab1 <- expression(paste("DIN (mg ", L^-1, ")"))
ylab2 <- expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")"))
ylab3 <- expression(paste("Si", O[2], " (mg ", L^-1, ")"))

# generic theme for all plots
ptheme <- theme_minimal() + 
  theme(
    axis.line.x = element_line(),
    axis.line.y = element_line(), 
    axis.title.x = element_blank(), 
    axis.text.x = element_blank()
  )

# colors
cols <- RColorBrewer::brewer.pal(9, 'Set1')[1:3]

## first set of plots
pdin1 <- prdnrmplot(toplo1$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  scale_y_continuous(ylab1) +
  theme(legend.position = 'top', legend.title = element_blank())
pleg1 <- g_legend(pdin1)
pdin1 <- pdin1 + theme(legend.position = 'none')

pdinsy1 <- seasyrplot(toplo1$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
  ptheme + 
  theme(legend.position = 'top', axis.title.y = element_blank())
pleg2 <- g_legend(pdinsy1)
pdinsy1 <- pdinsy1 + theme(legend.position = 'none')

pchl1 <- prdnrmplot(toplo1$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  theme(legend.position = 'none') + 
  scale_y_continuous(ylab2)

pchlsy1 <- seasyrplot(toplo1$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
  ptheme + 
  theme(legend.position = 'none', axis.title.y = element_blank())

psio1 <- prdnrmplot(toplo1$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
  scale_x_date('Year') +
  scale_y_continuous(ylab3)

psiosy1 <- seasyrplot(toplo1$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
  ptheme + 
  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
  scale_x_date('Day of year', date_labels = "%m/%d")

# Set the widths
pA1 <- ggplot_gtable(ggplot_build(pdin1))
pB1 <- ggplot_gtable(ggplot_build(pchl1))
pC1 <- ggplot_gtable(ggplot_build(psio1))
maxWidth = unit.pmax(pA1$widths[2:3], pB1$widths[2:3], pC1$widths[2:3])
pA1$widths[2:3] <- maxWidth
pB1$widths[2:3] <- maxWidth
pC1$widths[2:3] <- maxWidth

# Set the widths for seasyrplots
pAsy1 <- ggplot_gtable(ggplot_build(pdinsy1))
pBsy1 <- ggplot_gtable(ggplot_build(pchlsy1))
pCsy1 <- ggplot_gtable(ggplot_build(psiosy1))
maxWidth = unit.pmax(pAsy1$widths[2:3], pBsy1$widths[2:3], pCsy1$widths[2:3])
pAsy1$widths[2:3] <- maxWidth
pBsy1$widths[2:3] <- maxWidth
pCsy1$widths[2:3] <- maxWidth

## second set of plots
pdin2 <- prdnrmplot(toplo2$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  scale_y_continuous(ylab1) +
  theme(legend.position = 'none')

pdinsy2 <- seasyrplot(toplo2$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
  ptheme + 
  theme(legend.position = 'none', axis.title.y = element_blank())

pchl2 <- prdnrmplot(toplo2$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  theme(legend.position = 'none') + 
  scale_y_continuous(ylab2)

pchlsy2 <- seasyrplot(toplo2$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
  ptheme + 
  theme(legend.position = 'none', axis.title.y = element_blank())

psio2 <- prdnrmplot(toplo2$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
  ptheme + 
  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
  scale_x_date('Year') +
  scale_y_continuous(ylab3)

psiosy2 <- seasyrplot(toplo2$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
  ptheme + 
  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
  scale_x_date('Day of year', date_labels = "%m/%d")

# Set the widths
pA2 <- ggplot_gtable(ggplot_build(pdin2))
pB2 <- ggplot_gtable(ggplot_build(pchl2))
pC2 <- ggplot_gtable(ggplot_build(psio2))
maxWidth = unit.pmax(pA2$widths[2:3], pB2$widths[2:3], pC2$widths[2:3])
pA2$widths[2:3] <- maxWidth
pB2$widths[2:3] <- maxWidth
pC2$widths[2:3] <- maxWidth

# Set the widths for seasyrplots
pAsy2 <- ggplot_gtable(ggplot_build(pdinsy2))
pBsy2 <- ggplot_gtable(ggplot_build(pchlsy2))
pCsy2 <- ggplot_gtable(ggplot_build(psiosy2))
maxWidth = unit.pmax(pAsy2$widths[2:3], pBsy2$widths[2:3], pCsy2$widths[2:3])
pAsy2$widths[2:3] <- maxWidth
pBsy2$widths[2:3] <- maxWidth
pCsy2$widths[2:3] <- maxWidth

grid.arrange(
  arrangeGrob(
    ncol = 1, heights = c(0.05, 1),
    arrangeGrob(pleg1, pleg2, ncol = 2),
    arrangeGrob(
      arrangeGrob(
        textGrob('(a) Station D7, Suisun Bay'), heights = c(0.1, 1), 
        arrangeGrob(
          arrangeGrob(pA1, pB1, pC1, ncol = 1, heights = c(0.85, 0.85, 1)),
          arrangeGrob(pAsy1, pBsy1, pCsy1, ncol = 1, heights = c(0.85, 0.85, 1)),
          ncol = 2
        )
      ),
      arrangeGrob(
        textGrob('(b) Station C10, Delta'), heights = c(0.1, 1),
        arrangeGrob(
          arrangeGrob(pA2, pB2, pC2, ncol = 1, heights = c(0.85, 0.85, 1)),
          arrangeGrob(pAsy2, pBsy2, pCsy2, ncol = 1, heights = c(0.85, 0.85, 1)),
          ncol = 2
        )
      ),
      ncol = 1 
    )
  )
)

@@

cap.val<-'Summaries of flow-normalized trends in chlorophyll ($\\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods. Summaries are means and percent changes based on annual means within the time periods.  Increasing values are in bold-italics.'

### 2016-07-13 sftrends_manu.tex
c10 d7 trends plot

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Wed Jun 22 07:22:19 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

@@

% trends in sio and chl at d7
%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:chlsio")%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods. Summaries are means and percent changes based on annual means within the time periods.  Increasing values are in bold-italics.\label{tab:chlsio}} 
\begin{center}
\begin{tabular}{lllcll}
\hline\hline
\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
\cline{2-3} \cline{5-6}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
\hline
{\bfseries All}&&&&&\tabularnewline
~~1976-2013&3.5&-80.1&&12.5&\textit{\textbf{33.6}}\tabularnewline
\hline
{\bfseries Annual}&&&&&\tabularnewline
~~1976-1985&7.8&-62.3&&10.5&\textit{\textbf{8.9}}\tabularnewline
~~1986-1994&2.2&-62.1&&12.4&\textit{\textbf{15.6}}\tabularnewline
~~1995-2003&1.6&\textit{\textbf{20.2}}&&13.7&\textit{\textbf{4.5}}\tabularnewline
~~2004-2013&2&\textit{\textbf{31.3}}&&13.6&-2\tabularnewline
\hline
{\bfseries Seasonal}&&&&&\tabularnewline
~~JFM&1.9&-43.2&&14.8&\textit{\textbf{7.3}}\tabularnewline
~~AMJ&4.3&-72.8&&12.9&\textit{\textbf{25.7}}\tabularnewline
~~JAS&4.8&-88.4&&11.1&\textit{\textbf{70.8}}\tabularnewline
~~OND&2.1&-79.4&&11.7&\textit{\textbf{41.7}}\tabularnewline
\hline
\end{tabular}\end{center}

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Thu Jun 23 14:00:35 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

{\centering \includegraphics[width=\maxwidth]{figs/d7c10trnds-1}

\caption{Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\it\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).  Covariation between nutrients, chl-{\it\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.  Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
\end{figure}

@@

% trends in sio and chl at d7
%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:chlsio")%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in chlorophyll ($\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods. Summaries are means and percent changes based on annual means within the time periods.  Increasing values are in bold-italics.\label{tab:chlsio}} 
\begin{center}
\begin{tabular}{lllcll}
\hline\hline
\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
\cline{2-3} \cline{5-6}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
\hline
{\bfseries All}&&&&&\tabularnewline
~~1976-2013&3.5&-80.1&&12.5&\textit{\textbf{33.6}}\tabularnewline
\hline
{\bfseries Annual}&&&&&\tabularnewline
~~1976-1985&7.8&-62.3&&10.5&\textit{\textbf{8.9}}\tabularnewline
~~1986-1994&2.2&-62.1&&12.4&\textit{\textbf{15.6}}\tabularnewline
~~1995-2003&1.6&\textit{\textbf{20.2}}&&13.7&\textit{\textbf{4.5}}\tabularnewline
~~2004-2013&2&\textit{\textbf{31.3}}&&13.6&-2\tabularnewline
\hline
{\bfseries Seasonal}&&&&&\tabularnewline
~~JFM&1.9&-43.2&&14.8&\textit{\textbf{7.3}}\tabularnewline
~~AMJ&4.3&-72.8&&12.9&\textit{\textbf{25.7}}\tabularnewline
~~JAS&4.8&-88.4&&11.1&\textit{\textbf{70.8}}\tabularnewline
~~OND&2.1&-79.4&&11.7&\textit{\textbf{41.7}}\tabularnewline
\hline
\end{tabular}\end{center}
